Bayesian integration of networks without gold standards

### 2011-09-11 src/main/latex/paper/paper.tex
More related work in introduction

\section{Results:}
We present a novel algorithm for computing the probability of experimental observations to be correct that operates without gold standard reference data. We show that our algorithm outperforms existing gold standard based methods. Finally, we apply the new algorithm on a large collection of genetic interaction and protein-protein interaction experiments.

@@

A popular integration paradigm is the construction of functional networks \citep{von_mering_string:_2003,warde-farley_genemania_2010}\note{Also cite Katherine}. Functional networks represent different types of relationships between biological entities in an abstract manner. Associations such as genetic interactions, protein-protein interactions, gene regulation and co-expression are combined into simple abstract statements of functional relatedness, which are termed functional interactions.

An alternative paradigm is semantic data integration \citep{cheung_yeasthub:_2005,smith_obo_2007,koehler_graph-based_2006}. These approaches aim at representing the biological information (and as much of its meaning as possible) in a computationally accessible fashion. Rather than generalizing over all types of associations between entities to infer functional interactions, each type of association is considered separately.

An important question in data integration is how to assess the degree of confidence in each statement, that is, how likely the statement is to be correct. Several popular solutions to this problem exist in the context of functional networks \citep{lee_probabilistic_2004,troyanskaya_bayesian_2003}.
These methods assess the quality of each input dataset against one or more additional datasets of higher quality, usually manually-curated collections. Based on the confidence measures gained from this comparison it is then possible to calculate a confidence measure for each functional interaction. The high-quality datasets used in these comparisons are often referred to as `gold standards'.

Inferring confidence assessments for semantically integrated datasets is challenging, because each single type of association must be scored separately. However, gold standards only exist for some types of associations. Therefore, it is desirable to find a method that can infer confidence measures on biological networks, preferrably in the absence of a gold standard, that is, based only on the existing experimental data. So far, only partial solutions to this problem exist. These solutions infer confidence measures for a specific type of association only, for example, protein-protein interactions \citep{bader_gaining_2004,braun_experimentally_2009,venkatesan_empirical_2009}.

%Whether or not this description is justified is a matter of dispute \REF.

%It would be desirable to investigate these ideas at a more detailed level, that is, moving on from the question: ``How likely are two given proteins to be functionally related?'', to more specific questions such as: ``How likely is it that two given proteins physically interact?''. However, at this level it becomes infeasible to rely on gold standards, \note{since the level of completeness and quality required has not yet been achieved.} simply because none exist \REF. Therefore, it is necessary to develop a method that can infer confidence measures for semantic statements about molecular biological entities, based only on the existing data.

>>>>

\section{Results:}
We present a novel algorithm for computing the probability of experimental observations to be correct that operates without gold standard reference data. We show that our algorithm outperforms existing gold standard-based methods. Finally, we apply the new algorithm to a large collection of genetic interaction and protein-protein interaction experiments.

@@

A popular integration paradigm is the construction of functional networks \citep{lee_probabilistic_2004,von_mering_string:_2003,james_integration_2009}. Functional networks represent different types of relationships between biological entities in an abstract manner. Associations such as genetic interactions, protein-protein interactions, gene regulation and co-expression are combined into simple abstract statements of functional relatedness, which are termed functional interactions.

An alternative paradigm is semantic data integration \citep{cheung_yeasthub:_2005,smith_obo_2007,koehler_graph-based_2006}\note{Cite Pathway commons here}. These approaches aim at representing the biological information (and as much of its meaning as possible) in a computationally accessible fashion. Rather than generalizing over all types of associations between entities to infer functional interactions, each type of association is considered separately.

An important question regarding such networks is how to assess the degree of confidence in each statement, that is, how likely the statement is to be correct. Several popular solutions to this problem exist for functional networks \citep{lee_probabilistic_2004}\note{REF Lycett when it's accepted}.
These methods assess the quality of each input dataset against one or more additional datasets of higher quality, usually manually-curated collections. Based on the confidence measures gained from this comparison it is then possible to calculate a confidence measure for each functional interaction. The high-quality datasets used in these comparisons are often referred to as `gold standards'.

The method described by Lee and colleagues evaluates each evidential dataset against such a gold standard and obtain a log likelihood score (LLS). Subsequently, for each interaction in question, a weighted sum is formed over the LLS scores of those datasets that report the interaction. The weights are chosen in a manner that represents the degree of dependency between the datasets \citep{lee_probabilistic_2004}.

James and colleagues describe a method that extends the original method of Lee and colleagues \note{Cite Sam here as soon as the paper is out}. Not only one, but several different gold standards are used to generate LLS scores for the datasets. Furthermore, instead of creating a score for each interaction via the weighted sum described above, this method computes an existence probability from the original LLS scores and then averages over the different probabilites according to the different gold standards. The authors show that any bias inherent in the used gold standards can thus be overcome. \note{REF Lycett paper again here}

These methods work very well for functional networks. However, inferring confidence assessments for semantic networks, rather than functional networks, is more challenging, because each single type of association must be scored separately. Reliable gold standards only exist for some of these types. The methods discussed above are thus only of limited use for assessing semantic networks. While solutions for specific types of data do exist, for example, protein-protein interactions \citep{troyanskaya_bayesian_2003,bader_gaining_2004,braun_experimentally_2009,venkatesan_empirical_2009}, These are again dependent on additional data. It would desirable to find a method that can infer confidence measures on biological networks in the absence of a gold standard, that is, based only on the existing experimental data.
